---
title: "MPG Predictor — Shiny App Pitch"
author: "Alejandro Dominguez"
date: "April 17, 2026"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, comment = "")
```

## The Problem

Fuel efficiency matters — for your wallet and the planet.

But how do **weight, horsepower, cylinders, and transmission** actually affect a
car's miles per gallon?

**MPG Predictor** is a Shiny app that lets anyone explore this relationship
interactively, with **zero** statistics knowledge required.

> **Try it live:** [MPG Predictor on shinyapps.io](https://your-account.shinyapps.io/ShinyApp)

## The Model Behind the App

We fit a **multiple linear regression** on the classic `mtcars` dataset (32 cars):

```{r model, echo=TRUE}
model <- lm(mpg ~ wt + hp + factor(cyl) + am, data = mtcars)
round(coef(model), 3)
```

The model captures **`r paste0(round(summary(model)$r.squared * 100, 1), "%")`** of the variance in MPG (R² = `r round(summary(model)$r.squared, 3)`).

## How Accurate Is It?

```{r accuracy-plot, echo=FALSE, fig.height=4, fig.width=7, fig.align='center'}
par(mar = c(4, 4, 2, 1))
predicted <- predict(model)
plot(mtcars$mpg, predicted,
     xlab = "Actual MPG", ylab = "Predicted MPG",
     main = "Predicted vs. Actual MPG", pch = 19, col = "#2980b9", cex = 1.3)
abline(0, 1, col = "#e74c3c", lwd = 2, lty = 2)
```

Points hug the 45° line — the model generalizes well across the dataset.
RMSE = **`r round(sqrt(mean((mtcars$mpg - predicted)^2)), 2)`** MPG.

## App Features at a Glance

| Feature | Details |
|---|---|
| **Weight slider** | 1.5 – 5.5 thousand lbs |
| **Horsepower slider** | 50 – 350 hp |
| **Cylinder selector** | Radio buttons: 4, 6, or 8 |
| **Transmission picker** | Automatic / Manual dropdown |
| **Live prediction** | Updates instantly as inputs change |
| **Comparison plot** | Your car plotted against real mtcars data |
| **Model details toggle** | Optional full regression summary |

Built-in documentation is displayed directly in the sidebar — no external links needed.

## Thank You — Try It Yourself!

**GitHub repo:**
[github.com/your-username/datasciencecourse/ShinyApp](https://github.com)

**Live app:**
[shinyapps.io link](https://your-account.shinyapps.io/ShinyApp)

### Quick Start

1. Move the **Weight** and **Horsepower** sliders
2. Pick **Cylinders** and **Transmission**
3. See your predicted MPG update in real time

*Built with R `r getRversion()`, Shiny, and presented with R Markdown (ioslides).*
