---
title: "Iris Flower Classifier — Shiny App Pitch"
author: "Alejandro Dominguez"
date: "April 17, 2026"
output:
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, comment = "")
library(class)
```

## Why an Iris Classifier?

The **iris dataset** is a cornerstone of machine learning — 150 flowers, 3 species,
4 measurements. But interacting with it in a console is not intuitive.

**This Shiny app** lets anyone — even without R experience — classify a flower by
simply moving sliders:

- Adjust **sepal length, sepal width, petal length, and petal width**
- Pick the **number of neighbors (k)** for the kNN algorithm
- Get an **instant prediction** with confidence score and visual comparison

## The Algorithm: k-Nearest Neighbors

```{r knn-demo}
test_flower <- data.frame(Sepal.Length = 5.8, Sepal.Width = 3.0,
                          Petal.Length = 4.0, Petal.Width = 1.2)
pred <- knn(iris[,1:4], test_flower, iris$Species, k = 5, prob = TRUE)
cat("Predicted species:", as.character(pred),
    "\nConfidence:", round(attr(pred, "prob") * 100), "%")
```

The model computes Euclidean distances from the new flower to all `r nrow(iris)` training points, then takes a majority vote of the **k** closest neighbors.

## How Separable Are the Species?

```{r species-plot, echo=FALSE, fig.height=4.2, fig.width=7, fig.align='center'}
par(mar = c(4, 4, 2, 1))
cols <- c(setosa = "#e74c3c", versicolor = "#3498db", virginica = "#2ecc71")
plot(iris$Petal.Length, iris$Petal.Width, col = cols[iris$Species],
     pch = 19, cex = 1.3, xlab = "Petal Length (cm)",
     ylab = "Petal Width (cm)", main = "Iris Petal Dimensions by Species")
legend("topleft", legend = names(cols), col = cols, pch = 19, pt.cex = 1.3)
```

Petal dimensions alone separate the three species well — **setosa** is clearly distinct, while **versicolor** and **virginica** overlap slightly.

## Try It Yourself

| Feature | Description |
|---|---|
| **4 measurement sliders** | Sepal & petal length/width |
| **k selector** | Tune model sensitivity (1–15) |
| **Live prediction + confidence** | Updates instantly |
| **Scatter plot** | Your flower plotted among 150 real flowers |
| **Neighbors table** | See the k closest real flowers |
| **Documentation tab** | Full instructions for new users |

**Live app:** [shinyapps.io/IrisClassifier](https://alexdo10.shinyapps.io/ShinyApp)
&nbsp;|&nbsp; **Source code:** [GitHub](https://github.com/AlexDO10/datasciencecourse/tree/main/ShinyApp)
