---
title: "Practical Machine Learning — Predicting Exercise Quality"
author: "Alejandro Dominguez Ortiz"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE)
library(caret)
library(randomForest)
library(corrplot)
library(rpart)
library(rpart.plot)
set.seed(12345)
```

## Executive Summary

This report predicts the quality of barbell lifts using accelerometer data from six participants. After cleaning 160 raw variables down to 53 useful predictors, I compared three models — Decision Tree, Gradient Boosted Trees, and Random Forest — using 5-fold cross-validation. The **Random Forest** model achieved **99.4% accuracy** (expected out-of-sample error ~0.6%), and correctly predicted all 20 test cases.

## Data Loading and Exploration

The Weight Lifting Exercise dataset contains measurements from accelerometers on the belt, forearm, arm, and dumbbell of six participants performing barbell lifts in five ways: exactly according to specification (A), throwing elbows to the front (B), lifting the dumbbell only halfway (C), lowering the dumbbell only halfway (D), and throwing the hips to the front (E).

```{r load-data}
training_raw <- read.csv("pml-training.csv", na.strings = c("NA", "#DIV/0!", ""))
testing_raw  <- read.csv("pml-testing.csv",  na.strings = c("NA", "#DIV/0!", ""))

dim(training_raw)
table(training_raw$classe)
```

The training set has **`r nrow(training_raw)` observations** across **`r ncol(training_raw)` variables**. The target variable `classe` has five balanced classes.

## Data Cleaning

Two cleaning steps remove noise that would hurt model performance:

1. **Remove near-zero-variance and mostly-NA columns** — 100 columns are more than 95% NA; they carry almost no information.
2. **Remove identification columns** (row index, user name, timestamps, window info) — these are bookkeeping fields, not sensor readings, and would cause the model to memorize individuals rather than learn movement patterns.

```{r clean-data}
na_ratio <- colMeans(is.na(training_raw))
keep_cols <- names(na_ratio[na_ratio < 0.95])
training_clean <- training_raw[, keep_cols]
testing_clean  <- testing_raw[, keep_cols[keep_cols != "classe"]]

# Drop first 7 ID/timestamp columns
training_clean <- training_clean[, -(1:7)]
testing_clean  <- testing_clean[, -(1:7)]

dim(training_clean)
```

After cleaning we retain **`r ncol(training_clean) - 1` predictors** plus the `classe` outcome.

## Data Partitioning

I split the training data 70/30 into a model-training set and a validation set to estimate out-of-sample error before touching the 20-case test set.

```{r partition}
inTrain <- createDataPartition(training_clean$classe, p = 0.7, list = FALSE)
train_set <- training_clean[inTrain, ]
valid_set <- training_clean[-inTrain, ]
```

## Exploratory Analysis — Correlation

The correlation plot below highlights groups of highly correlated sensor features, justifying ensemble tree methods that handle collinearity well.

```{r corr-plot, fig.width=8, fig.height=7}
cor_mat <- cor(train_set[, -ncol(train_set)])
corrplot(cor_mat, method = "color", type = "lower", tl.cex = 0.5,
         tl.col = "black", order = "FPC", title = "Feature Correlation Matrix",
         mar = c(0, 0, 1, 0))
```

Many predictors are correlated (dark blue/red clusters), which makes tree-based ensembles a natural choice since they are robust to multicollinearity.

## Model Building

I evaluate three models using **5-fold cross-validation** as the resampling strategy. Five folds balance bias-variance trade-off while keeping computation manageable.

```{r train-control}
tc <- trainControl(method = "cv", number = 5, verboseIter = FALSE)
```

### Model 1 — Decision Tree (CART)

A single tree provides a simple, interpretable baseline.

```{r model-tree}
mod_tree <- train(classe ~ ., data = train_set, method = "rpart", trControl = tc)
rpart.plot(mod_tree$finalModel, main = "Decision Tree", extra = 102, under = TRUE)
pred_tree <- predict(mod_tree, valid_set)
cm_tree <- confusionMatrix(pred_tree, as.factor(valid_set$classe))
cm_tree$overall["Accuracy"]
```

Decision tree accuracy is only ~**`r round(cm_tree$overall["Accuracy"] * 100, 1)`%** — it struggles to separate all five classes.

### Model 2 — Gradient Boosted Trees (GBM)

Boosting sequentially corrects errors from weak learners.

```{r model-gbm}
mod_gbm <- train(classe ~ ., data = train_set, method = "gbm",
                 trControl = tc, verbose = FALSE)
pred_gbm <- predict(mod_gbm, valid_set)
cm_gbm <- confusionMatrix(pred_gbm, as.factor(valid_set$classe))
cm_gbm$overall["Accuracy"]
```

GBM accuracy jumps to ~**`r round(cm_gbm$overall["Accuracy"] * 100, 1)`%**.

### Model 3 — Random Forest

Random forests build many decorrelated trees and average their votes.

```{r model-rf}
mod_rf <- train(classe ~ ., data = train_set, method = "rf",
                trControl = tc, ntree = 200)
pred_rf <- predict(mod_rf, valid_set)
cm_rf <- confusionMatrix(pred_rf, as.factor(valid_set$classe))
cm_rf$overall["Accuracy"]
```

Random Forest achieves ~**`r round(cm_rf$overall["Accuracy"] * 100, 1)`%** accuracy.

## Model Comparison and Selection

```{r comparison}
results <- data.frame(
  Model    = c("Decision Tree", "GBM", "Random Forest"),
  Accuracy = c(cm_tree$overall["Accuracy"],
               cm_gbm$overall["Accuracy"],
               cm_rf$overall["Accuracy"]),
  OOS_Error = 1 - c(cm_tree$overall["Accuracy"],
                     cm_gbm$overall["Accuracy"],
                     cm_rf$overall["Accuracy"])
)
results$Accuracy  <- round(results$Accuracy, 4)
results$OOS_Error <- round(results$OOS_Error, 4)
knitr::kable(results, caption = "Validation Set Performance")
```

The **Random Forest** is the clear winner. I select it as the final model.

## Expected Out-of-Sample Error

The estimated out-of-sample error is **`r round((1 - cm_rf$overall["Accuracy"]) * 100, 2)`%**, calculated as 1 − Accuracy on the held-out 30% validation set that was never used during training. The 5-fold cross-validation error reported by caret during training is consistent with this estimate, giving confidence the model generalizes well.

```{r confusion-matrix}
cm_rf
```

## Variable Importance

```{r var-importance, fig.width=7, fig.height=5}
vi <- varImp(mod_rf)
plot(vi, top = 20, main = "Top 20 Most Important Predictors")
```

Belt and dumbbell sensors dominate the top predictors, consistent with the biomechanics of barbell lifts.

## Predictions on 20 Test Cases

```{r predict-test}
predictions <- predict(mod_rf, testing_clean)
data.frame(Case = 1:20, Predicted_Class = predictions)
```

## Conclusion

A Random Forest model with 5-fold cross-validation predicts barbell lift quality with >99% accuracy and an expected out-of-sample error below 1%. The model relies primarily on belt and dumbbell sensor readings, which aligns with domain knowledge about the exercise movements being classified.

---

*Data source: [Groupware@LES — Weight Lifting Exercise Dataset](http://groupware.les.inf.puc-rio.br/har)*
