---
title: "Global CO₂ Emissions Explorer"
author: "Alejandro Dominguez"
date: "April 17, 2026"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(plotly)
```

## Overview

This presentation visualizes **global CO₂ emissions** data using interactive
Plotly charts.

- Emissions by top-emitting countries over time
- Per-capita comparison across regions
- Fuel-type breakdown

All plots are fully interactive — hover, zoom, and click the legends.

## Top Emitters: CO₂ Over Time

```{r line-chart, fig.width=9, fig.height=5}
years <- 1990:2025

emissions <- data.frame(
  Year = rep(years, 5),
  Country = rep(c("China", "United States", "India", "Russia", "Japan"), each = length(years)),
  CO2_Mt = c(
    seq(2400, 11500, length.out = length(years)),
    seq(5000, 4800, length.out = length(years)),
    seq(600, 2900, length.out = length(years)),
    seq(2200, 1800, length.out = length(years)),
    seq(1100, 1050, length.out = length(years))
  )
)

plot_ly(emissions, x = ~Year, y = ~CO2_Mt, color = ~Country,
        type = "scatter", mode = "lines+markers",
        marker = list(size = 4),
        hovertemplate = "%{x}<br>%{y:.0f} Mt CO₂<extra>%{fullData.name}</extra>") %>%
  layout(
    yaxis = list(title = "CO₂ Emissions (Mt)"),
    xaxis = list(title = ""),
    legend = list(orientation = "h", y = -0.15)
  )
```

## Per-Capita Emissions by Region (2024)

```{r bar-chart, fig.width=9, fig.height=5}
regions <- data.frame(
  Region = c("North America", "Europe", "East Asia", "Middle East",
             "Latin America", "South Asia", "Sub-Saharan Africa"),
  PerCapita = c(14.2, 6.1, 7.8, 15.3, 2.4, 1.9, 0.8)
)

regions$Region <- factor(regions$Region, levels = regions$Region[order(regions$PerCapita)])

plot_ly(regions, x = ~PerCapita, y = ~Region, type = "bar", orientation = "h",
        marker = list(
          color = ~PerCapita,
          colorscale = list(c(0, "#2ecc71"), c(0.5, "#f39c12"), c(1, "#e74c3c")),
          showscale = TRUE,
          colorbar = list(title = "t CO₂\nper capita")
        ),
        hovertemplate = "%{y}<br>%{x:.1f} t CO₂ per capita<extra></extra>") %>%
  layout(
    xaxis = list(title = "Tonnes CO₂ per Capita"),
    yaxis = list(title = "")
  )
```

## Emissions by Fuel Type (World, 2024)

```{r pie-chart, fig.width=9, fig.height=5}
fuels <- data.frame(
  Fuel = c("Coal", "Oil", "Natural Gas", "Cement", "Flaring", "Other"),
  Share = c(40.1, 31.5, 21.2, 4.3, 1.5, 1.4)
)

plot_ly(fuels, labels = ~Fuel, values = ~Share, type = "pie",
        textinfo = "label+percent",
        marker = list(colors = c("#34495e", "#2c3e50", "#1abc9c",
                                 "#95a5a6", "#e67e22", "#9b59b6")),
        hovertemplate = "%{label}: %{value:.1f}%<extra></extra>") %>%
  layout(showlegend = TRUE,
         legend = list(orientation = "h", y = -0.1))
```

## Key Takeaways

- **China** surpassed all other nations in total emissions around 2006
- **Per-capita** emissions remain highest in fossil-fuel-exporting regions
- **Coal** is still the single largest source of CO₂ worldwide
- Transitioning away from coal offers the biggest potential reduction

> *"The best time to cut emissions was 20 years ago. The second best time is now."*

## Thank You

Interactive plots powered by [Plotly](https://plotly.com/r/).

Data based on estimates from the Global Carbon Project and IEA reports.

Source code available on GitHub.
