---
title: "Impact of Severe Weather Events on Public Health and the Economy in the United States"
author: "Alejandro Dominguez"
date: "April 17, 2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE)
```

## Synopsis

This report analyzes the U.S. National Oceanic and Atmospheric Administration's
(NOAA) storm database to determine which types of severe weather events are most
harmful to population health and which have the greatest economic consequences.
The dataset spans from 1950 to November 2011 and contains 902,297 records.
To assess public health impact, we aggregate fatalities and injuries by event
type. For economic impact, we compute total property and crop damage in dollars
by converting the damage exponent codes (K, M, B) to their numeric multipliers.
Our analysis reveals that **tornadoes** are by far the most dangerous event for
population health, causing the most fatalities and injuries combined. In terms
of economic damage, **floods** cause the greatest total property and crop losses,
followed by hurricanes/typhoons and tornadoes. These findings can help emergency
managers prioritize resources for the most destructive event categories.

## Data Processing

We load the raw compressed data file directly (`.csv.bz2`) and select only the
columns relevant to our two research questions.

```{r load-data}
storm <- read.csv("StormData.csv.bz2")
dim(storm)
```

The dataset contains `r nrow(storm)` observations and `r ncol(storm)` variables.
We keep only the columns needed for our analysis: event type, fatalities,
injuries, property damage, and crop damage.

```{r subset}
cols <- c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP",
          "CROPDMG", "CROPDMGEXP")
storm <- storm[, cols]
str(storm)
```

### Processing Economic Damage

The property and crop damage values are split across two columns each: a numeric
value (`PROPDMG`, `CROPDMG`) and an exponent code (`PROPDMGEXP`, `CROPDMGEXP`).
The exponent codes represent multipliers: **K** = thousands, **M** = millions,
**B** = billions. We convert these to full dollar amounts.

```{r damage-multiplier}
get_multiplier <- function(exp) {
    exp <- toupper(exp)
    mult <- rep(1, length(exp))
    mult[exp == "K"] <- 1e3
    mult[exp == "M"] <- 1e6
    mult[exp == "B"] <- 1e9
    mult[exp == "H"] <- 1e2
    mult[exp %in% as.character(0:9)] <- 10^as.numeric(exp[exp %in% as.character(0:9)])
    mult
}

storm$PROP_TOTAL <- storm$PROPDMG * get_multiplier(storm$PROPDMGEXP)
storm$CROP_TOTAL <- storm$CROPDMG * get_multiplier(storm$CROPDMGEXP)
storm$ECON_TOTAL <- storm$PROP_TOTAL + storm$CROP_TOTAL
```

### Aggregating by Event Type

We aggregate fatalities, injuries, and economic damage by event type.

```{r aggregate}
health <- aggregate(cbind(FATALITIES, INJURIES) ~ EVTYPE, data = storm, sum)
health$HEALTH_TOTAL <- health$FATALITIES + health$INJURIES
health <- health[order(-health$HEALTH_TOTAL), ]

econ <- aggregate(cbind(PROP_TOTAL, CROP_TOTAL, ECON_TOTAL) ~ EVTYPE,
                  data = storm, sum)
econ <- econ[order(-econ$ECON_TOTAL), ]
```

## Results

### Which types of events are most harmful to population health?

We examine the top 10 event types by combined fatalities and injuries.

```{r health-table}
top_health <- head(health, 10)
top_health[, c("EVTYPE", "FATALITIES", "INJURIES", "HEALTH_TOTAL")]
```

**Figure 1** shows the top 10 weather event types by total health impact
(fatalities + injuries). Tornadoes dominate overwhelmingly, causing over five
times more harm than the next event type.

```{r figure1, fig.width=10, fig.height=6, fig.cap="Figure 1: Top 10 weather events most harmful to population health (1950–2011)."}
par(mar = c(8, 6, 3, 1))
colors <- c("#e74c3c", rep("#3498db", 9))
bp <- barplot(top_health$HEALTH_TOTAL,
              names.arg = top_health$EVTYPE,
              las = 2, col = colors, border = NA,
              main = "Top 10 Weather Events Most Harmful to Population Health",
              ylab = "Total Fatalities + Injuries")
text(bp, top_health$HEALTH_TOTAL, labels = format(top_health$HEALTH_TOTAL, big.mark = ","),
     pos = 3, cex = 0.8)
```

To understand the breakdown between fatalities and injuries, **Figure 2** shows
both components side by side for the top 10 events.

```{r figure2, fig.width=10, fig.height=6, fig.cap="Figure 2: Fatalities vs. injuries for the top 10 most harmful weather events."}
par(mar = c(8, 6, 3, 6))
mat <- t(as.matrix(top_health[, c("FATALITIES", "INJURIES")]))
colnames(mat) <- top_health$EVTYPE
barplot(mat, beside = TRUE, las = 2, col = c("#e74c3c", "#f39c12"), border = NA,
        main = "Fatalities vs. Injuries by Weather Event Type",
        ylab = "Count")
legend("topright", legend = c("Fatalities", "Injuries"),
       fill = c("#e74c3c", "#f39c12"), border = NA)
```

Tornadoes cause the most injuries by a wide margin. Excessive heat ranks second
for fatalities specifically, highlighting that lethal events are not always the
ones causing the most injuries.

### Which types of events have the greatest economic consequences?

We examine the top 10 event types by total economic damage (property + crop).

```{r econ-table}
top_econ <- head(econ, 10)
top_econ$PROP_BILLIONS <- round(top_econ$PROP_TOTAL / 1e9, 2)
top_econ$CROP_BILLIONS <- round(top_econ$CROP_TOTAL / 1e9, 2)
top_econ$TOTAL_BILLIONS <- round(top_econ$ECON_TOTAL / 1e9, 2)
top_econ[, c("EVTYPE", "PROP_BILLIONS", "CROP_BILLIONS", "TOTAL_BILLIONS")]
```

**Figure 3** shows the top 10 event types by total economic damage, split into
property and crop components.

```{r figure3, fig.width=10, fig.height=6, fig.cap="Figure 3: Top 10 weather events with the greatest economic consequences (1950–2011). Damage in billions of USD."}
par(mar = c(8, 6, 3, 6))
mat_econ <- t(as.matrix(top_econ[, c("PROP_TOTAL", "CROP_TOTAL")])) / 1e9
colnames(mat_econ) <- top_econ$EVTYPE
barplot(mat_econ, beside = FALSE, las = 2, col = c("#2980b9", "#2ecc71"), border = NA,
        main = "Top 10 Weather Events by Economic Damage",
        ylab = "Total Damage (Billions USD)")
legend("topright", legend = c("Property Damage", "Crop Damage"),
       fill = c("#2980b9", "#2ecc71"), border = NA)
```

Floods cause the greatest total economic damage at over $150 billion,
predominantly from property losses. Hurricanes/typhoons rank second, also
driven mainly by property damage. Drought, while not among the top property
damage events, causes substantial crop losses.

### Summary of Key Findings

| Rank | Most Harmful to Health | Greatest Economic Damage |
|------|------------------------|--------------------------|
| 1 | Tornado | Flood |
| 2 | Excessive Heat | Hurricane/Typhoon |
| 3 | TSTM Wind | Tornado |
| 4 | Flood | Storm Surge |
| 5 | Lightning | Hail |

Tornadoes are the most dangerous weather events for population health, while
floods impose the heaviest economic burden. Decision-makers should prioritize
tornado preparedness for life safety and flood mitigation for economic
protection.
